Naturalness and Long-Lived Beyond the Standard Model Particles

### 2017-08-02 tex/draft.tex
full draft of viability

@@

        \noindent \begin{figure}[htbp] \begin{center}
            \includegraphics[width=0.5\textwidth]{images/pixel_old_new}
            \caption{Diagram showing barrel pixel layers before and after the winter 2016/17 upgrade.\fxnote{cite website found through Andrew's citations}}
            \label{pixel_old_new}
        \end{center} \end{figure}

@@

        \noindent \begin{figure}[htbp] \begin{center}
        \includegraphics[width=0.6\textwidth]{images/material_map}
            \caption{Map of material dense regions of the ATLAS detector.\fxnote{explain more}~\cite{atals_displaced}}
        \label{material_map}
        \end{center} \end{figure}

@@

        \fxnote{ugh, this is too hard. just steal diagram from paper or make in jaxodraw. Maybe even use simulated event instead, but make sure I can draw these if asked?}  
        \noindent \begin{figure}[htbp] \begin{center}
        \begin{subfigure}[htbp]{0.3\textwidth} \begin{center}
        \includegraphics[width=\textwidth]{images/disappearing_ew_process}
        \end{center} \end{subfigure}
        \qquad
        \begin{subfigure}[htbp]{0.3\textwidth} \begin{center}
        \includegraphics[width=\textwidth]{images/disappearing_qcd_process}
        \end{center} \end{subfigure}
        \fxnote{Can't get it to work in tikz-feynman. Just take from paper for now.} 
        %\feynmandiagram [layered layout] {
        %    i1 -- [plain] a [blob],
        %    i2 -- [plain] a,
        %    a -- [plain, boson] b [blob],
        %    a -- [plain] f1,
        %    a -- [plain, boson] f2,
        %    b -- [plain, boson] f3,
        %    b -- [scalar] f4,
        %};
            \caption{Processes\fxnote{explain}.~\cite{atlas_disappearing}}
        \label{disappearing_processes}
        \end{center} \end{figure}

@@

    \subsection{Stopped Particles}
        \fxnote{Add calorimeter search?}  
        \fxnote{cite CMS PAS EXO-17-004}
        \fxnote{mention how stopped particles complement HSCP search?}
        \fxnote{too redundant with overview?}
        With long enough lifetimes and sufficient mass, LLPs could come to rest in the detector before decaying. The eventual decay would then lead to energy deposits in the calorimeters or hits in the muon system that could be out of time with proton bunch crossings. A recent CMS analysis looked for such a signal in dimuon final states using the full 2015 and 2016 datasets.

@@

        \fxnote{check if tof, beta, or bx is better} 
        \noindent \begin{figure}[htbp] \begin{center}
        \includegraphics[width=\textwidth]{images/stopped_reject_cosmic_tof}
        \caption{Cosmic muon rejection with MS.\fxnote{cite analysis}}
        \label{stopped_muon_differentiation}
        \end{center} \end{figure}

@@

        \fxnote{ok to skip mchamp? Be able to explain why} 
        \noindent \begin{figure}[htbp] \begin{center}
        \begin{subfigure}[htbp]{0.5\textwidth} \begin{center}
        \includegraphics[width=\textwidth]{images/stopped_limit_gluino_cross_section}
        \end{center} \end{subfigure}
        \begin{subfigure}[htbp]{0.45\textwidth} \begin{center}
        \includegraphics[width=\textwidth]{images/stopped_limit_gluino_mass}
        \end{center} \end{subfigure}
        \caption{Limits.\fxnote{cite analysis}}
        \label{stopped_limits}
        \end{center} \end{figure}

    \subsection{Heavy Stable Charged Particles}
        \fxnote{cite CMS EXO-15-010}
        \fxnote{nice dE/dx plot (fig 1)}
        Particles that live for more than a couple ns\fxnote{how do I typeset standalone units?} \fxnote{Check this and be more exact} will appear stable in CMS and ATLAS because they can traverse the entire detector before decaying. If such a particle is also sufficiently massive, it can be catigorized as a heavy stable charged particle (HSCP). HSCPs are characterterized by a greater rate of energy loss via ionization due to the lower speed with which they pass through detector material\fxnote{redundant w/ overview. Probably better explained here than there}. Additionally, HSCPs can be fractionally, singly, or multiply charged, which also affects their rate of energy loss, as shown in Fig.~\ref{hscp_dedx}.

        \noindent \begin{figure}[htbp] \begin{center}
        \includegraphics[width=0.5\textwidth]{images/hscp_dedx}
            \caption{Ionization energy loss.\fxnote{cite analysis}}
        \label{hscp_dedx}
        \end{center} \end{figure}

@@

        \fxnote{I show seperate trk, trk+tof limits but never explain that in text} 
        \noindent \begin{figure}[htbp] \begin{center}
        \begin{subfigure}[htbp]{0.45\textwidth} \begin{center}
        \includegraphics[width=\textwidth]{images/hscp_limit_tracker_only}
        \end{center} \end{subfigure}
        \begin{subfigure}[htbp]{0.45\textwidth} \begin{center}
        \includegraphics[width=\textwidth]{images/hscp_limit_tracker_tof}
        \end{center} \end{subfigure}
        \caption{Limits.\fxnote{cite analysis}}
        \label{hscp_limits}
        \end{center} \end{figure}

\section{Experimental Viability of Natural SUSY}
    \fxnote{Andrew says: This should really be your most important section. You should try to bring together the results of the searches you talked about (and of other long-lived particle searches and even of more general SUSY searches) and explain what they mean for a possible natural explanation of EWSB. At what point can we claim that there is no natural solution, or at least that there is no natural solution in terms of SUSY? Essentially, nobody cares about what you talked about in Section 6; what's important is that you can give these results a meaningful context in the bigger picture of a natural explanation of EWSB.}

    Interpreting current limits in the context of naturalness requires some set of criteria that determine if a model is natural. No well-defined, universally-agreed-upon criteria exist, but certain themes recur in many discussions of natural model building\fxnote{better be able to confidently cite several sources}. The following discussion lays out a simplistic set of criteria for natural SUSY models and attempts to assess the status of natural SUSY in light of current experimental limits\fxnote{Did I just steal Chris's langauge?}.

    When building a natural SUSY model, certain constraints are nearly unavoidable\fxnote{Dangerous. What if Linda knows how to avoid them?}. Specifically, the Higgsino\fxnote{careful, there could be multiple Higgsinos. Do I mean 'lightest Higgsino'?}, which corrects $m_H$ at tree level; the stop\fxnote{mention that stop is more important than other sfermions b/c large yukawa coupling? I'll have to clarify that somewhere}, which corrects $m_H$ at one loop; and the gluino, which corrects mStop at one loop and therefore $m_H$ at two, must all be light enough to avoid excessive fine tuning.\fxnote{Need diagrams and correction terms for sure} Furthermore, the stop and gluino masses are correlated --- a large difference between the two would itself be unnatural.

    \fxnote{cite cornering} 
    \fxnote{define \mu, Q?} 
    \fxnote{make sure these are logically and notationally consistent w/ earlier higgs corrections} 
    \noindent \begin{equation}
        \delta \mHiggs^2 = \mu^2
        \label{higgsino_correction}
    \end{equation}
    \noindent \begin{equation}
        \delta \mHiggs^2 \approx - \frac{3}{8\pi^2} y_t^2 m_{\tilde{t}}^2 \ln{\left(\frac{\Lambda}{Q}\right)}
        \label{stop_correction}
    \end{equation}
    \noindent \begin{equation}
        \delta \mHiggs^2 \approx - \frac{g_3^2 y_t^2}{4\pi^4} M_3^2 \ln{\left(\frac{\Lambda}{Q}\right)}^2
        \label{gluino_correction}
    \end{equation}

@@

    \fxnote{cite cornering} 
    \noindent \begin{equation}
        \Delta_i = \frac{\partial \ln{\mHiggs^2}}{\partial \ln{\alpha_i}} \quad ,
        \label{fine_tuning}
    \end{equation}

    \noindent where $\alpha_i$ is a particular model parameter and $\Delta_i$ is the degree of fine tuning from variations in that parameter. The overall degree of fine tuning for the model can then be assigned as $\max{\left(\Delta_i\right)}$.

    Determining the acceptable degree of fine tuning is even more open ended. Before null results and serious constraints started rolling in, it was common\fxnote{cite something} to refer to the $\Delta \lesssim 10$ regions of parameter space as natural. As models are constrained, physicists seem willing to accept greater degrees of fine tuning. Instead of asking, ``what degree of fine tuning should be considered natural?", Feng proposes physicists ask themselves, ``If SUSY were discovered with such-and-such degree of fine tuning, would I consider the naturalness problem solved and move on to different research questions?" Anecdotally, Feng reports that acceptable levels of fine tuning tend to increase when the question is posed in this way~\cite{feng}.

    So, how is natural SUSY fairing? Fig.~\ref{atlas_susy_summary} summarizes limits on many SUSY models and topologies from ATLAS searches. CMS limits are generally similar\fxnote{make sure this is true}, but a recent plot was not available.\fxnote{give general summary}

    \noindent \begin{figure}[htbp] \begin{center}
    \includegraphics[width=0.8\textwidth]{images/atlas_susy_summary}
    \caption{Summary of current experimental limits on SUSY models from ATLAS searches.\fxnote{cite something}}
    \label{atlas_susy_summary}
    \end{center} \end{figure}

    Each of the limits in Fig.~\ref{atlas_susy_summary} comes with a particular set of assumptions, making it difficult to draw general conclusions. Some authors try to solve this problem by recasting analyses in terms of a few general models that illustrate much of the natural SUSY landscape. Fig.~\ref{cornering_limits} does just this with several searches using $\approx \SI{15}{\femto\barn^{-1}}$ of \SI{13}{\tera\electronvolt} data for the cases of Effective SUSY, RPV SUSY, Hidden Valley SUSY, and a model that combines RPV with Effective SUSY. The authors find that for $\mu \leq \SI{100}{\giga\electronvolt}$ and $\Lambda = \SI{20}{\tera\electronvolt}$, a small region of $\Delta < 10$ parameter space remains viable in all cases.\fxnote{mention that HL-LHC should rule everything out?}

        \noindent \begin{figure}[htbp] \begin{center}
        \begin{subfigure}[htbp]{0.4\textwidth} \begin{center}
        \includegraphics[width=\textwidth]{images/cornering_effective_100}
        \end{center} \end{subfigure}
        \begin{subfigure}[htbp]{0.4\textwidth} \begin{center}
        \includegraphics[width=\textwidth]{images/cornering_rpv_100}
        \end{center} \end{subfigure}
        \begin{subfigure}[htbp]{0.4\textwidth} \begin{center}
        \includegraphics[width=\textwidth]{images/cornering_hv_100}
        \end{center} \end{subfigure}
        \begin{subfigure}[htbp]{0.4\textwidth} \begin{center}
        \includegraphics[width=\textwidth]{images/cornering_rpv_effective_100}
        \end{center} \end{subfigure}
        \caption{Limits.\fxnote{cite cornering}}
        \label{cornering_limits}
        \end{center} \end{figure}

    As viable parameter space for natural SUSY continues to shrink, it is important to identify and fill in the gaps left by conventional analyses. LLP searches such as those in Sec.~\ref{analyses} do just that by opening up sensitivity to lifetimes that would otherwise be missed. Fig.~\ref{fate_limits} shows stop mass limits from several Run I LLP searches that have been recast using a simple RPV model\fxnote{cite fate}. By combining results from several analyses, the authors constrain the stop mass for a broad and continuous range of lifetimes. LLP searches are slowly excluding the natural parameter space

    Despite increasing contraints, SUSY remains one of the least-tuned options for natural BSM physics\fxnote{Make sure I can explain this and make sure it isn't plagarized from Cornering}.\fxnote{Add exotica summary plot to show how constrained everything else is?}

\fxnote{When should symbols be italicized and when should they be mathrm'd?} 
\fxnote{fine-tuning or fine tuning?}

>>>>

% Things to check
\fxnote{When should symbols be italicized and when should they be mathrm'd?} 
\fxnote{fine-tuning or fine tuning?}

@@

        \noindent \begin{figure}[htbp] \begin{center}
            \includegraphics[width=0.4\textwidth]{images/pixel_old_new}
            \caption{Diagram showing barrel pixel layers before and after the winter 2016/17 upgrade.\fxnote{cite website found through Andrew's citations}}
            \label{pixel_old_new}
        \end{center} \end{figure}

@@

        \noindent \begin{figure}[htbp] \begin{center}
        \includegraphics[width=0.6\textwidth]{images/material_map}
            \caption{Map of material dense regions of the ATLAS detector.\fxnote{explain more}~\cite{atlas_displaced}}
        \label{material_map}
        \end{center} \end{figure}

@@

        \fxnote{ugh, this is too hard. just steal diagram from paper or make in jaxodraw. Maybe even use simulated event instead, but make sure I can draw these if asked?}  
        \noindent \begin{figure}[htbp] \begin{center}
        \begin{subfigure}[htbp]{0.2\textwidth} \begin{center}
        \includegraphics[width=\textwidth]{images/disappearing_ew_process}
        \end{center} \end{subfigure}
        \qquad
        \begin{subfigure}[htbp]{0.2\textwidth} \begin{center}
        \includegraphics[width=\textwidth]{images/disappearing_qcd_process}
        \end{center} \end{subfigure}
        \fxnote{Can't get it to work in tikz-feynman. Just take from paper for now.} 
        %\feynmandiagram [layered layout] {
        %    i1 -- [plain] a [blob],
        %    i2 -- [plain] a,
        %    a -- [plain, boson] b [blob],
        %    a -- [plain] f1,
        %    a -- [plain, boson] f2,
        %    b -- [plain, boson] f3,
        %    b -- [scalar] f4,
        %};
            \caption{Processes\fxnote{explain}.~\cite{atlas_disappearing}}
        \label{disappearing_processes}
        \end{center} \end{figure}

@@

    \subsection{Stopped Particles}
        \fxnote{Add calorimeter search?}  
        \fxnote{too redundant with overview?}
        With long enough lifetimes and sufficient mass, LLPs could come to rest in the detector before decaying. The eventual decay would then lead to energy deposits in the calorimeters or hits in the muon system that could be out of time with proton bunch crossings. A recent CMS analysis looked for such a signal in dimuon final states using the full 2015 and 2016 datasets~\cite{cms_stopped}.

@@

        \fxnote{check if tof, beta, or bx is better} 
        \noindent \begin{figure}[htbp] \begin{center}
        \includegraphics[width=\textwidth]{images/stopped_reject_cosmic_tof}
            \caption{Cosmic muon rejection using time-of-flight information from the muon system.~\cite{cms_stopped}}
        \label{stopped_muon_differentiation}
        \end{center} \end{figure}

@@

        \fxnote{ok to skip mchamp? Be able to explain why} 
        \noindent \begin{figure}[htbp] \begin{center}
        \begin{subfigure}[htbp]{0.5\textwidth} \begin{center}
        \includegraphics[width=\textwidth]{images/stopped_limit_gluino_cross_section}
        \end{center} \end{subfigure}
        \begin{subfigure}[htbp]{0.45\textwidth} \begin{center}
        \includegraphics[width=\textwidth]{images/stopped_limit_gluino_mass}
        \end{center} \end{subfigure}
            \caption{Upper limits on production cross section as a function of gluino lifetime (left) and mass (right)~\cite{cms_stopped}}
        \label{stopped_limits}
        \end{center} \end{figure}

    \subsection{Heavy Stable Charged Particles}
        \fxnote{nice dE/dx plot (fig 1)}
        Particles that live for more than a couple \SI{}{ns}\fxnote{Check this and be more exact} will appear stable in CMS and ATLAS because they can traverse the entire detector before decaying. If such a particle is also sufficiently massive, it can be catigorized as a heavy stable charged particle (HSCP). HSCPs are characterterized by a greater rate of energy loss via ionization due to the lower speed with which they pass through detector material\fxnote{redundant w/ overview. Probably better explained here than there}. Additionally, HSCPs can be fractionally, singly, or multiply charged, which also affects their rate of energy loss, as shown in Fig.~\ref{hscp_dedx}~\cite{cms_hscp}.

        \noindent \begin{figure}[htbp] \begin{center}
        \includegraphics[width=0.5\textwidth]{images/hscp_dedx}
            \caption{Ionization energy loss as a function of momentum for paticles of varying mass and charge~\cite{cms_hscp}}
        \label{hscp_dedx}
        \end{center} \end{figure}

@@

        \fxnote{I show seperate trk, trk+tof limits but never explain that in text} 
        \noindent \begin{figure}[htbp] \begin{center}
        \begin{subfigure}[htbp]{0.45\textwidth} \begin{center}
        \includegraphics[width=\textwidth]{images/hscp_limit_tracker_only}
        \end{center} \end{subfigure}
        \begin{subfigure}[htbp]{0.45\textwidth} \begin{center}
        \includegraphics[width=\textwidth]{images/hscp_limit_tracker_tof}
        \end{center} \end{subfigure}
        \caption{Production cross section limits on several particles as a function of mass for the tracker-only and tracker+TOF analyses~\cite{cms_hscp}}
        \label{hscp_limits}
        \end{center} \end{figure}

\section{Experimental Viability of Natural SUSY}
    Interpreting current limits in the context of naturalness requires some set of criteria that determine if a model is natural. No well-defined, universally-agreed-upon criteria exist, but certain themes recur in many discussions of natural model building~\cite{craig, vestiges, evans_toward_full, feng, cornering, kim}. The following discussion lays out a simplistic set of criteria for natural SUSY models and attempts to assess the status of natural SUSY in light of current experimental limits\fxnote{Did I just steal Chris's langauge?}.

    \fxnote{I should be able to combine or reorganize the next two paragraphs to flow better} 
    The parameters of interest are the Higgsino\fxnote{careful - do I mean lightest higgsino?}, stop, and gluino masses, which correct the Higgs mass at tree-level, one-loop, and two-loops, respectively (see Eqs.~\ref{higgsino_correction}-~\ref{gluino_correction}). Furthermore, the stop and gluino masses are correlated --- a large difference between the two would itself be unnatural~\cite{cornering}.

    \fxnote{define \mu, Q?} 
    \fxnote{make sure these are logically and notationally consistent w/ earlier higgs corrections} 
    \noindent \begin{equation}
        \delta \mHiggs^2 = \mu^2
        \label{higgsino_correction}
    \end{equation}
    \noindent \begin{equation}
        \delta \mHiggs^2 \approx - \frac{3}{8\pi^2} y_t^2 m_{\tilde{t}}^2 \ln{\left(\frac{\Lambda}{Q}\right)}
        \label{stop_correction}
    \end{equation}
    \noindent \begin{equation}
        \delta \mHiggs^2 \approx - \frac{g_3^2 y_t^2}{4\pi^4} M_3^2 \ln{\left(\frac{\Lambda}{Q}\right)}^2
        \label{gluino_correction}
    \end{equation}

@@

    \noindent \begin{equation}
        \Delta_i = \frac{\partial \ln{\mHiggs^2}}{\partial \ln{\alpha_i}} \quad ,
        \label{fine_tuning}
    \end{equation}

    \noindent where $\alpha_i$ is a particular model parameter and $\Delta_i$ is the degree of fine tuning assoicated with that parameter. The overall degree of fine tuning for the model can then be assigned as $\max{\left(\Delta_i\right)}$~\cite{cornering}.

    Determining the acceptable degree of fine tuning is even more open ended. Before null results and serious constraints started rolling in, it was common\fxnote{feng} to refer to the $\Delta \lesssim 10$ regions of parameter space as natural, but as models are constrained, physicists seem willing to accept greater degrees of fine tuning. Feng proposes an interesting way to recast this problem: instead of asking which regions of parameter space are most natural, physicists should instead ask themselves, ``If superpartners were discovered, what level of fine fine-tuning would be sufficient to convince me that the gauge hierarchy problem is solved by sypersymmetry and I should move on to researhing other problems?"\fxnote{paraphrase for brevity?} Anecdotally, acceptable levels of fine tuning tend to increase when the question is posed in this way~\cite{feng}.

    Exact upper limits on natural sparticle masses are therefore hard to come by. A reasonable and somewhat typical set of values is given in~\cite{drees_kim}: $m_{\tilde{H}} \lesssim \SI{500}{\giga\electronvolt}$, $m_{\tilde{t}} \lesssim \SI{1.5}{\tera\electronvolt}$, and $m_{\tilde{g}} \lesssim \SI{3}{\tera\electronvolt}$. Fig.~\ref{atlas_susy_summary} summarizes current SUSY limits from ATLAS searches (CMS limits are generally similar, but a recent plot was not available). Upon inspection, one sees that experimental limits are starting to push on naturalness bounds in many scenarios.

    \fxnote{what do I cite?}  
    \noindent \begin{figure}[htbp] \begin{center}
    \includegraphics[width=0.8\textwidth]{images/atlas_susy_summary}
    \caption{Summary of current experimental limits on SUSY models from ATLAS searches.}
    \label{atlas_susy_summary}
    \end{center} \end{figure}

    Each of the limits in Fig.~\ref{atlas_susy_summary} comes with a particular set of assumptions, making it difficult to draw general conclusions. Some authors try to solve this problem by recasting analyses in terms of a few general models that illustrate much of the natural SUSY landscape. Fig.~\ref{cornering_limits} shows the results from one such attempt using $\approx \SI{15}{\femto\barn^{-1}}$ of \SI{13}{\tera\electronvolt} data for the cases of Effective SUSY, RPV SUSY, Hidden Valley SUSY, and a model that combines RPV with Effective SUSY. The authors find that for $\mu \leq \SI{100}{\giga\electronvolt}$ and $\Lambda = \SI{20}{\tera\electronvolt}$, a small region of $\Delta < 10$ parameter space remains viable in all cases but project that all considered $\Delta < 10$ parameter space will be excluded within the next few years\fxnote{check}.

        \noindent \begin{figure}[htbp] \begin{center}
        \begin{subfigure}[htbp]{0.3\textwidth} \begin{center}
        \includegraphics[width=\textwidth]{images/cornering_effective_100}
        \end{center} \end{subfigure}
        \begin{subfigure}[htbp]{0.3\textwidth} \begin{center}
        \includegraphics[width=\textwidth]{images/cornering_rpv_100}
        \end{center} \end{subfigure}
        \begin{subfigure}[htbp]{0.3\textwidth} \begin{center}
        \includegraphics[width=\textwidth]{images/cornering_hv_100}
        \end{center} \end{subfigure}
        \begin{subfigure}[htbp]{0.3\textwidth} \begin{center}
        \includegraphics[width=\textwidth]{images/cornering_rpv_effective_100}
        \end{center} \end{subfigure}
        \caption{Limits.~\cite{cornering}}
        \label{cornering_limits}
        \end{center} \end{figure}

    As viable parameter space for natural SUSY continues to shrink, it is important to identify and fill in the gaps left by conventional analyses. LLP searches such as those in Sec.~\ref{analyses} do just that by opening up sensitivity to lifetimes that would otherwise be missed. Fig.~\ref{fate_limits}, for example, shows stop mass limits from several Run I LLP searches that have been recast using a simple RPV model~\cite{fate}. By combining results from several analyses, the authors constrain the stop mass for a broad and continuous range of lifetimes that would have been invisible to conventional searches.

        \noindent \begin{figure}[htbp] \begin{center}
        \includegraphics[width=\textwidth]{images/fate_limits}
        \caption{Limits.~\cite{fate}}
        \label{fate_limits}
        \end{center} \end{figure}

    CMS and ATLAS should be able to exclude or discover a wide range of natural SUSY models over the next several years~\cite{cornering}. In the absence of a discovery, physicists will have to look long and hard at the role of naturalness in the SM. In the meantime, however, SUSY remains one of the least-tuned options for natural BSM physics\fxnote{double check, then cite cornering, feng, nate, and/or giudice}\fxnote{Should I include exo summary plot to show how constrained compositeness and extra dimensions are?}. In the final years of the hunt for natrual SUSY, CMS and ATLAS will need to apply a combination of conventional and LLP searches in order to cover natural parameter space.
\fxnote{need to answer, "which searches are important?}
